UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-05-29 ms/main.tex
some edits

The HRS forms several large groupings and clusters, reflecting both ethnicity (figure~\ref{fig:supp_umap_hrs_eth}) and admixture proportions (figure~\ref{fig:umap_hrs_admix}). Gradients in admixture proportion are visible within the predominantly Hispanic cluster, but not within the predominantly Black cluster, perhaps because the variance in ancestry proportions is greater among Hispanics. The "White Not Hispanic" (WNH) group forms several interconnected clusters, and these do not correspond to broad geographical areas (figure~\ref{fig:supp_hrs_born}). The clarity of the interconnected clusters varies by parameterization, but they consistently form a large, roughly connected group.

@@

The largest body in figure~\ref{fig:umap_ukbb_pc10} consists of the White British and Irish populations. The Irish population concentrates in one part of this group, but many individuals are also scattered throughout the British-identified population. Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters. Chinese individuals form a cluster, within a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well. The East Asian and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group". The patchwork of genetic neighbourhoods is connected by trails of admixed individuals, which converge at a nexus of individuals with a variety of ethnicities. Many claim mixed ancestry, and there are clusters of individuals who belong to an "other ethnic group". Using data on countries of birth, we identified many groups in figure~\ref{fig:supp_ukbb_cob}, and confirmed they appeared in intuitive areas with, e.g., Japanese and Filipino groups being projected near Chinese groups.

Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection in figure~\ref{fig:umap_ukbb_pc10} coloured in by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly. Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe-wide data\cite{novembre2008europe}. Most admixture lines have coordinates in the south east of the UK, corresponding to the city of London and reflecting its high migrant population. The detailed shape of extended clusters is not stable as we vary the number of PCs included. Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB. The shape of the White British cluster is different, and we observe finer patterns of geographic variation, yet the observations made earlier are maintained.

@@

The HRS contains genotype data of 12,454 American individuals across all 50 states who have provided racial identity (10,434 White, 1,652 Black, 368 Other) as well as whether they identify as Hispanic (1,203 total) and, if so, whether they identify as Mexican-American (705 total)\cite{juster1995overview}. We crossed these three variables to form a composite self-reported ethnicity resulting in 10 categories (e.g. White Hispanic Mexican-American), and considered birth regions based on the 10 census regions and divisions used by the US Census Bureau. Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian/Native American, and European populations using RFMIX \cite{Maples:2013fia}. We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.

>>>>

The HRS forms several large groupings and clusters, reflecting both ethnicity (figure~\ref{fig:supp_umap_hrs_eth}) and admixture proportions (figure~\ref{fig:umap_hrs_admix}). Gradients in admixture proportion are visible within the predominantly Hispanic cluster, but not within the predominantly Black cluster, perhaps because the variance in ancestry proportions is greater among Hispanics. The "White Not Hispanic" (WNH) group forms several interconnected clusters, and these do not correspond to broad geographical areas (figure~\ref{fig:supp_hrs_born}). The clarity of the interconnected clusters varies by parameterization, but they consistently form a large, roughly connected group. \sgcomment{Last sentence is a bit vague. Remove?}

@@

The largest body in figure~\ref{fig:umap_ukbb_pc10} consists of the White British and Irish populations. The Irish population concentrates in one part of this group, but many individuals are also scattered throughout the British-identified population. Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters. Chinese individuals form a cluster, within a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well. The East Asian and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group". The patchwork of genetic neighbourhoods is connected by trails of admixed individuals, which converge at a nexus of individuals with a variety of ethnicities. Many claim mixed ancestry, and there are clusters of individuals who belong to an "other ethnic group". Using data on countries of birth, we identified many finer groups in figure~\ref{fig:supp_ukbb_cob}, and confirmed they appeared in intuitive areas with, e.g., Japanese and Filipino groups being projected near Chinese groups.

Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from figure~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly. Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe-wide data\cite{novembre2008europe}. Most admixture lines have coordinates in the south east of the UK, where the city of London isa located, likely reflecting its high migrant population. The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved. Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB. %The shape of the White British cluster is different, and we observe finer patterns of geographic variation, yet the observations made earlier are maintained.

@@

The HRS contains genotype data of 12,454 American individuals across all 50 states who have provided racial identity (10,434 White, 1,652 Black, 368 Other) as well as whether they identify as Hispanic (1,203 total) and, if so, whether they identify as Mexican-American (705 total)\cite{juster1995overview}. We crossed these three variables to form a composite self-reported ethnicity resulting in 10 categories (e.g. White Hispanic Mexican-American), and considered birth regions based on the 10 census regions and divisions used by the US Census Bureau. Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian, and European populations using RFMIX \cite{Maples:2013fia}. We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.
